\null\vfill\small
\noindent\emph{\thejournal}.~\thejournalsubtitle. \\ 
\noindent Volume~\thevolume: \thevolumetitle. \\ \par
\noindent ISSN: 1879-6095 \\
\noindent ISSN-L: 1573-3084 \\ \par
\noindent\doclicenseIcon \ \theyear \\ \par 
\noindent This volume, including all its textual contents, is licensed under a \doclicenseLongNameRef \ license, and made available in Open Access at \href{\thewebsite}{\thewebsite}. The authors of the individual contributions, who are identified as such, retain the copyright over their original work.  All authors who included third-party images in their contributions confirmed that they had permission to do so. In such cases, the original copyright and licenses continue to apply when removed from the context of the contribution they were included in. \\ \par
\noindent For more information on the \doclicenseNameRef license, please refer to: \\ \doclicenseURL.  \\ \par

\noindent Cover image: Authorial additions and revisions (\textasciitilde 1918) in: \textit{Maironis, Pavasario Balsai Ir Kame Išganymas}. 4th ed., Kaunas, 1913, pp. 10–11, Maironis Lithuanian Literature Museum, No. GEK 102103. The printed text and handwritten additions and revisions in this image are by Maironis, whose works have since entered the public domain. The spread was digitized by the Maironis Lithuanian Literature Museum as commissioned by Paulius V. Subačius, and the resulting digital image generously released by the Museum without further copyright restrictions. When reusing this image, we kindly request users to respect the Museum's wishes, and mention the place of handling in full where appropriate. The image features in this volume's contribution by Paulius V. Subačius (\pageref{fig:subacius:maironis1918}; Figure \ref{fig:subacius:maironis1918}). \\ \par

\noindent This volume was typeset in \LaTeX~by Wout Dillen using \href{https://github.com/WoutDLN/variantx}{varian\TeX} ~--- a reusable template for journals in the Humanities, also developed by Wout Dillen. \mbox{varian\TeX} ~is open source, available on GitHub, and deposited in the \href{https://about.zenodo.org}{Zenodo Open Science Repository}. DOI: \href{https://zenodo.org/record/3484652#.X0PdDy2w3kI}{10.5281/zenodo.3484652}.
\newpage